An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2011-01-26 paper.tex
New references for kD trees, some additions and minor edits.

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
These neighborhoods should be non-overlapping and fill parameter
space.  \ilya{I don't think this is generic enough: for example, Littenberg and Cornish also constructed an interpolation, but one in which multiple points could be in the same "neighborhood", while other neighborhoods could be empty. I.e., the algorithm below is one way of solving this problem, but not an alternative statement of it.}
To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF. \ilya{This last statement is presumably only true if the size of the neighborhood associated with each point is roughly the inverse of the local point density.  I think it's worthwhile to point that out that your kD trees allow you to just that in a computationally efficient manner.}  In fact, the proposed
jumps are drawn from a piecewise-constant (i.e.\ constant on each
neighborhood) interpolation of the PDF.  There are various techniques
that could be used to construct the set of neighborhoods associated
with each point.

@@

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions associated with each MCMC sample.  The
Voronoi region associated with a sample contains all the parameter
space points that are closer to that sample than any other sample.
The Voronoi region decomposition into neighborhoods is, in a sense,
maximally adaptive, in contrast to the approach of
Ref.~\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi regions requires a metric on
parameter space, which may be difficult or impossible to define.  A
larger difficulty with the Voronoi approach is that the computational
cost for computing the Voronoi regions for $N$ points increases
rapidly with dimension, $d$, as $N^X$ \will{Find X in the literature;
  add references.  I think it's $X = d/2$}.

We propose to use a decomposition of parameter space into
neighborhoods based on a datastructure called a kD-tree
\will{References.}.  The decomposition is more adaptive than the cells
of Ref.~\cite{Littenberg2009}, and more efficient in high dimension
than the Voronoi decomposition.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{kdtree.png}
  \end{center}
  \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
    constructed around a set of points that are normally distributed
    about the origin in two dimensions.  As the points become denser
    around the origin, the typical neighborhood gets smaller.  The
    interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
    where $N$ is the total number of points (and boxes).}
\end{figure}

In order to use the kD-tree interpolation in an MCMC, we must be able
to quickly find the neighborhood associated with a given point.  This
can be accomplished in $\order{\log N}$ time and constant space with
the following algorithm.  Given the point, $\vtheta_i$ --- which could
either be a point in the chain of MCMC samples if we are trying to
propose a jump, or a previously proposed jump if we are trying to
compute the jump probability distribution (see
Eq.~\ref{eq:p-accept}) --- and the tree, $T$:
\begin{enumerate}
\item If $T$ contains exactly one sample point, then its box is the
  associated neighborhood.  Otherwise:
\item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
  contained in the ``left'' sub-tree, then return to Step 1,
  considering this sub-tree; otherwise return to Step 1, considering
  the ``right'' sub-tree.
\end{enumerate}

>>>>

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
We choose these neighborhoods to be non-overlapping and fill parameter
space.  To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF.  In other words, the size
of a neighborhood is inversely proportional to the local point
density.  The proposed jumps are drawn from a piecewise-constant
(i.e.\ constant on each neighborhood) interpolation of the PDF.  There
are various techniques that could be used to construct the set of
neighborhoods associated with each point.

@@

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions associated with each MCMC sample.  The
Voronoi region associated with a sample contains all the parameter
space points that are closer to that sample than any other sample.
The Voronoi region decomposition into neighborhoods is, in a sense,
maximally adaptive, in contrast to the approach of
Ref.~\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi regions requires a metric on
parameter space, which may be difficult or impossible to define.  A
larger difficulty with the Voronoi approach is that the computational
cost for computing the Voronoi regions increases rapidly with
dimension.  % Apparently the exact computational cost is not known in
            % arbitrary dimension.

We propose to use a decomposition of parameter space into
neighborhoods based on a datastructure called a kD-tree (see, e.g.\
\cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
efficient in high dimension than the Voronoi decomposition.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{kdtree.png}
  \end{center}
  \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
    constructed around a set of points that are normally distributed
    about the origin in two dimensions.  As the points become denser
    around the origin, the typical neighborhood gets smaller.  The
    interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
    where $N$ is the total number of points (which is also the number
    of boxes).}
\end{figure}

In order to use the kD-tree interpolation as a jump proposal in an
MCMC, we must be able to quickly find the neighborhood associated with
a given point to compute the jump probability (see
Eq.~\ref{eq:p-accept}).  This can be accomplished in $\order{\log N}$
time and constant space with the following algorithm, which is a
standard binary tree search.  Given the point, $\vtheta_i$ and the
tree, $T$:
\begin{enumerate}
\item If $T$ contains exactly one sample point, then its box is the
  associated neighborhood.  Otherwise:
\item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
  contained in the ``left'' sub-tree, then return to Step 1,
  considering this sub-tree; otherwise return to Step 1, considering
  the ``right'' sub-tree.
\end{enumerate}
